The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2013-02-27 script/en/chapters/numbertheory.tex
CT Script: Added 48th M-Prime and some minor changes (DE and EN)

This is not the case, however, for 
modular multiplication.
\begin{theorem}\label{thm-zth-multinv}
A natural number $a$ from the set $\{1, \cdots, m-1\}$ has one inverse if
and only if this number and the modulo $m$ are co-prime\footnote{Two whole
numbers $a$ and $b$ are co-prime\index{Number!co-prime} if and only if
${\rm gcd}(a, b) = 1$.\\
If $p$ is prime and $a$ is a random whole number that is not a multiple of
$p$, then $p$ and a are co-prime.\\
Further name to the topic co-prime (with $a_i \in \mathbb{Z}, i=1, \cdots, n$):
\begin{enumerate}
\item $a_1,a_2, \cdots, a_n$ are {\em relatively prime}
\index{Prime number!relative prime}\index{Relative prime},
if $ \mbox{gcd}(a_1, \cdots , a_n) =1.$
\item An even stronger request for more than two numbers is :\\
     $a_1, \cdots , a_n$ are {\em in pairs relatively prime}, if for all $i=1, \cdots, n$ and 
$j=1, \cdots , n$ with $ i \neq j $: $ \mbox{gcd} (a_i, a_j) =1. $
\end{enumerate}
\begin{example}{:}\\
$2,3,6 $ are relatively prime, because $ \mbox{gcd} (2,3,6)=1.$
\end{example}
They are not in pairs relatively prime, because
$ \mbox{gcd} (2,6)=2>1.$}\index{gcd}, in other words if $a$ and $m$ have 
no common prime factors.
\end{theorem}
Since $m=5$ is prime, the numbers $1$ to $4$ are relatively prime to $5$ and {\bf each} of these numbers has a 
multiplicative inverse in mod $5$.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                   & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17 & 18 \\
\hline
\hline  
$5*i$                & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80  & 85 & 90   \\
remainder                & 5 & 10 & 15  & 3  & 8 & 13  & {\bf 1}  & 6 & 11 & 16  & 4  & 9 & 14  & 2  & 7 & 12 & 0  & 5   \\
\hline
$6*i$                 & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108   \\
remainder                & 6 & 12  & {\bf 1}  & 7 & 13  & 2  & 8 & 14  & 3  & 9 & 15  & 4 & 10 & 16  & 5 & 11 & 0  & 6   \\
\hline
\end{tabular}
\end{center} 
\caption{Multiplication table modulo $17$ (for $a=5$ and $a=6$)}\label{SrcArith1a}\label{mulmod17}
\end{table}
Between $i=1, \cdots, m$, all values between $0, \cdots, m-1$ occur for the remainders, because both $5$ and $6$ are 
also relatively prime\index{Prime number!relative prime}\index{Relative prime}
to the modulus $m=17$.
\enlargethispage{0.5cm}

@@

Solution: We have $e*d \equiv 1 {\rm ~(mod~} J(26))$.
\begin{itemize}
\item[] The reduced remainder set $R' = \mathbb{Z}_{26}^* = \{ 1, 3, 5, 7, 9, 11, 15, 17, 19, 21, 23, 25 \}$ are
the elements in $\mathbb{Z}_{26},$ which have a multiplicative inverse,
that is which are relatively prime\index{Prime number!relative prime}\index{Relative prime}
to $26$ (see \ref{def-zth-znmult}).
\item[] The reduced remainder set $R''$ contains only the elements of $R'$ that are relatively prime to 
        $J(n) = 12:  R'' = \{ 1, 5, 7, 11 \}$.
\item[] For every $e$ in $R''$ there exists a $d$ in $R''$ such that $a \equiv (a^e)^d {\rm ~(mod~} n)$.
\end{itemize}
For every $e$ in $R''$, there exists therefore precisely one element (not necessarily different from $e$) 
such that $e*d \equiv 1 {\rm ~(mod~} J(26))$.

@@

\vskip +10 pt
\begin{theorem}\label{thm-zth-ordp}
  Given a modulus $n$ and a number $a$, relative prime to $n$ the
  following holds: \\
  The set $\{ a^i\; (\mbox{mod }n) |\; i = 1,\dots,J(n)\}$ equals the
  multiplicative group $Z_n^*$ if and only if $ord_n(a) = J(n)$.\footnote{%
    For prime moduli $p$ all $a$ with $0 < a < p$ are of order $J(p) = p - 1$.
    Compare table~\ref{expmod45} for an example. In this case $a^i (\mbox{mod }n)$ goes
    through all the values $1,\dots,p-1$. Exhausting all possible values of
    the set is an important cryptographic proposition (compare theorem
    \ref{thm-zth-exhperm}). This determines a permutation\index{Permutation}
    $\pi(p-1)$.}$^,$\footnote{%
    Table~\ref{expmod46} demonstrates that for composite moduli $n$ not all $a$ are of
    maximal order $J(n)$. In this example only $5,7,11,15,17,19\mbox{ and }21$
    are of order 22.
  }
\end{theorem}
\noindent The multiplicative group $Z_n^*$ only contains all values from
$1$ to $n-1$, if n is prime (see \ref{def-zth-znmult}).

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|cccc|}
\hline
	& {\bf Decimal digits} & {\bf Binary digits} & {\bf Factored on} & {\bf Factored by} \\
\hline
	&&&& \\
	RSA-768 & 	232 & 768 & Dec 2010 & Thorsten Kleinjung et al. \\
	RSA-200 & 	200 & 663 & May 2005 & Jens Franke et al. \\
	RSA-640\footnotemark & 	193 & 640 & Nov 2005 & Jens Franke et al. \\
	RSA-576 & 	174 & 576 & Dec 2003 & Jens Franke et al. \\
	RSA-160 & 	160 & 530 & Apr 2003 & Jens Franke et al. \\
	RSA-155	&	155 & 512 & Aug 1999 & Herman te Riele et al. \\
	\dots &&&& \\
	C307 & 		307 & 1017 & May 2007 & Jens Franke et al. \\
	C176 & 		176 & 583 & May 2005 & Kazumaro Aoki et al. \\
	C158 & 		158 & 523 & Jan 2002 & Jens Franke et al. \\
\hline
\end{tabular}
\caption{The current factoring records (as of Nov. 2012)}    % Eyecatcher_neue-Mersenne 
\label{factorizationrecords}
\end{center}
\end{table} 
\footnotetext{%
A research group of the GISA solved this challenge which was awarded with
20,000 US dollar using the GNFS method. The researchers needed about five
months to divide this number into its both 320 bit long prime factors.

@@

$$ [RSA{-}768~number] ~~\longleftrightarrow{}~~ [47th ~known~Mersenne~prime] $$
$$ 232 ~~ \longleftrightarrow{} ~~ 43,112,609 ~~~~~$$
$$ [see~table~\ref{factorizationrecords}] ~~\longleftrightarrow{}~~ [see~table~\ref{L_n_Largest_Known-Primes}] ~~~~~~~~~~~~~~~~$$

@@

  \item[{\bf 2.}] Select an arbitrary $e \in \{2, \cdots, n-1\}$ such that: \\
                  $e$ is relatively prime to $J(n) = (p-1)*(q-1)$. \\
                  We can then ``throw away'' $p$ and $q$.

>>>>

This is not the case, however, for 
modular multiplication.
\begin{theorem}\label{thm-zth-multinv}
A natural number $a$ from the set $\{1, \cdots, m-1\}$ has one inverse if
and only if this number and the modulo $m$ are co-prime\footnote{Two whole
numbers $a$ and $b$ are co-prime\index{Number!co-prime} if and only if
${\rm gcd}(a, b) = 1$.\\
If $p$ is prime and $a$ is a random whole number that is not a multiple of
$p$, then $p$ and a are co-prime.\\
Further name to the topic co-prime (with $a_i \in \mathbb{Z}, i=1, \cdots, n$):
\begin{enumerate}
\item $a_1,a_2, \cdots, a_n$ are {\em relatively prime}
\index{Prime number!relative prime}\index{Number!relative prime},
if $ \mbox{gcd}(a_1, \cdots , a_n) =1.$
\item An even stronger request for more than two numbers is :\\
     $a_1, \cdots , a_n$ are {\em in pairs relatively prime}, if for all $i=1, \cdots, n$ and 
$j=1, \cdots , n$ with $ i \neq j $: $ \mbox{gcd} (a_i, a_j) =1. $
\end{enumerate}
\begin{example}{:}\\
$2,3,6 $ are relatively prime, because $ \mbox{gcd} (2,3,6)=1.$
\end{example}
They are not in pairs relatively prime, because
$ \mbox{gcd} (2,6)=2>1.$}\index{gcd}, in other words if $a$ and $m$ have 
no common prime factors.
\end{theorem}
Since $m=5$ is prime, the numbers $1$ to $4$ are relatively prime to $5$ and {\bf each} of these numbers has a 
multiplicative inverse in mod $5$.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                   & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17 & 18 \\
\hline
\hline  
$5*i$                & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80  & 85 & 90   \\
remainder                & 5 & 10 & 15  & 3  & 8 & 13  & {\bf 1}  & 6 & 11 & 16  & 4  & 9 & 14  & 2  & 7 & 12 & 0  & 5   \\
\hline
$6*i$                 & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108   \\
remainder                & 6 & 12  & {\bf 1}  & 7 & 13  & 2  & 8 & 14  & 3  & 9 & 15  & 4 & 10 & 16  & 5 & 11 & 0  & 6   \\
\hline
\end{tabular}
\end{center} 
\caption{Multiplication table modulo $17$ (for $a=5$ and $a=6$)}
\label{SrcArith1a}\label{mulmod17}
\end{table}

Between $i=1, \cdots, m$, all values between $0, \cdots, m-1$ occur for the remainders,
because both $5$ and $6$ are also relatively prime
\index{Prime number!relative prime}\index{Number!relative prime}
to the modulus $m=17$.
\enlargethispage{0.5cm}

@@

Solution: We have $e*d \equiv 1 {\rm ~(mod~} J(26))$.
\begin{itemize}
\item[] The reduced remainder set $R' = \mathbb{Z}_{26}^* = \{ 1, 3, 5, 7, 9, 11, 15, 17, 19, 21, 23, 25 \}$ are
the elements in $\mathbb{Z}_{26},$ which have a multiplicative inverse,
that is which are relatively prime
\index{Prime number!relative prime}\index{Number!relative prime}
to $26$ (see \ref{def-zth-znmult}).
\item[] The reduced remainder set $R''$ contains only the elements of $R'$ that are relatively prime to 
        $J(n) = 12:  R'' = \{ 1, 5, 7, 11 \}$.
\item[] For every $e$ in $R''$ there exists a $d$ in $R''$ such that $a \equiv (a^e)^d {\rm ~(mod~} n)$.
\end{itemize}
For every $e$ in $R''$, there exists therefore precisely one element (not necessarily different from $e$) 
such that $e*d \equiv 1 {\rm ~(mod~} J(26))$.

@@

\vskip +10 pt
\begin{theorem}\label{thm-zth-ordp}
  Given a modulus $n$ and a number $a$, relative prime\index{Prime number!relative prime}\index{Number!relative prime} to $n$, the following holds: \\
  The set $\{ a^i\; (\mbox{mod }n) |\; i = 1,\dots,J(n)\}$ equals the
  multiplicative group $Z_n^*$ if and only if $ord_n(a) = J(n)$.\footnote{%
    For prime moduli $p$ all $a$ with $0 < a < p$ are of order $J(p) = p - 1$.
    Compare table~\ref{expmod45} for an example. In this case $a^i (\mbox{mod }n)$ goes
    through all the values $1,\dots,p-1$. Exhausting all possible values of
    the set is an important cryptographic proposition (compare theorem
    \ref{thm-zth-exhperm}). This determines a permutation\index{Permutation}
    $\pi(p-1)$.}$^,$\footnote{%
    Table~\ref{expmod46} demonstrates that for composite moduli $n$ not all $a$ are of
    maximal order $J(n)$. In this example only $5,7,11,15,17,19\mbox{ and }21$
    are of order 22.
  }
\end{theorem}
\noindent The multiplicative group $Z_n^*$ only contains all values from
$1$ to $n-1$, if n is prime (see \ref{def-zth-znmult}).

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|cccc|}
\hline
	& {\bf Decimal digits} & {\bf Binary digits} & {\bf Factored on} & {\bf Factored by} \\
\hline
	&&&& \\
	RSA-768 & 	232 & 768 & Dec 2010 & Thorsten Kleinjung et al. \\
	RSA-200 & 	200 & 663 & May 2005 & Jens Franke et al. \\
	RSA-640\footnotemark & 	193 & 640 & Nov 2005 & Jens Franke et al. \\
	RSA-576 & 	174 & 576 & Dec 2003 & Jens Franke et al. \\
	RSA-160 & 	160 & 530 & Apr 2003 & Jens Franke et al. \\
	RSA-155	&	155 & 512 & Aug 1999 & Herman te Riele et al. \\
	\dots &&&& \\
	C307 & 		307 & 1017 & May 2007 & Jens Franke et al. \\
	C176 & 		176 & 583 & May 2005 & Kazumaro Aoki et al. \\
	C158 & 		158 & 523 & Jan 2002 & Jens Franke et al. \\
\hline
\end{tabular}
\caption{The current factoring records (as of Nov. 2012)}    % Eyecatcher
\label{factorizationrecords}
\end{center}
\end{table} 
\footnotetext{%
A research group of the GISA solved this challenge which was awarded with
20,000 US dollar using the GNFS method. The researchers needed about five
months to divide this number into its both 320 bit long prime factors.

@@

$$ [RSA{-}768~number] ~~\longleftrightarrow{}~~ [47th ~known~Mersenne~prime] $$
$$ 232 ~~ \longleftrightarrow{} ~~ 17,425,170 ~~~~~$$
$$ [see~table~\ref{factorizationrecords}] ~~\longleftrightarrow{}~~ [see~table~\ref{L_n_Largest_Known-Primes}] ~~~~~~~~~~~~~~~~$$

@@

  \item[{\bf 2.}] Select an arbitrary $e \in \{2, \cdots, n-1\}$ such that: \\
                  $e$ is relatively prime
                  \index{Prime number!relative prime}\index{Number!relative prime}
                  to $J(n) = (p-1)*(q-1)$. \\
                  We can then ``throw away'' $p$ and $q$.
